---
title: 'code'
author: "Natalie Alexander"
date: "2023-09-23"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
rm(list=ls())

```

# 1. Libraries

```{r, message=FALSE, echo=TRUE}

#load in libraries
library(tidyverse)
library(tidyr)
library(dplyr)
library(knitr)
library(magick)
library(repmod)
library(ggplot2)
library(corrplot)
library(openxlsx)
library(stats)
library(arules)
library(caret)
library(rpart)
library(e1071)
library(randomForest)
library(grid)
library("arulesViz")
```

# 2. Data

```{r}
#################################Read in Data###################################

#set path to access data files

##CHANGE PATH ACCORDING TO WHERE THE DATA FILE IS STORED ON YOUR COMPUTER!
path = "C:\\Users\\natal\\OneDrive - University of Cape Town\\UCT\\Unsupervised\\Assignments\\Zazo_Assignment1"

##CHANGE PATH ACCORDING TO WHERE THE TABLES AND FIGURES WILL BE SAVED ON YOUR COMPUTER!
path_to_figures_tables = "C:\\Users\\natal\\OneDrive - University of Cape Town\\UCT\\Unsupervised\\Assignments\\Zazo_Assignment1\\submit\\Question2\\figures_tables"

#read in data
cad_data = read.xlsx(paste0(path, "\\CAD dataset.xlsx"), sheet = 2)



```


# 3. Exploratory Data Analysis and Data Pre-processing

## 3.1. Inspect data

```{r}
############################Inspect data dimensions#############################

#check dimensions of data
##dimensions of data set:
dim(cad_data) #303 rows by 56 columns data matrix

##number of columns:
ncol(cad_data) #56 columns

##number of rows:
nrow(cad_data) #303 rows

##########################check for missing data################################
cad_data[!complete.cases(cad_data),]


##############################Inspect column names##############################

#column names
colnames(cad_data)

```

## 3.2. Head and tail of data

```{r}
##############################Inspect data #####################################

#check head of data
(head_of_cad_data = head(cad_data))

#save table data
repmod::make_word_table(head_of_cad_data, 
                        paste0(path_to_figures_tables, 
                               "\\head_of_cad_data"), 
                        info = NULL, use.rownames = TRUE)



#check tail of data
(tail_of_cad_data = tail(cad_data))

#save table data
repmod::make_word_table(tail_of_cad_data, 
                        paste0(path_to_figures_tables, 
                               "\\tail_of_cad_data"), 
                        info = NULL, use.rownames = TRUE)
```


## 3.3. Data types

```{r}
########################Check data types of data################################

str(cad_data)

########################Convert Yes/No classes to binary#########################

cad_data[,"Obesity"] = ifelse(cad_data[,"Obesity"] == "Y", 1, 0)
cad_data[,"CRF"] = ifelse(cad_data[,"CRF"] == "Y", 1, 0)
cad_data[,"CVA"] = ifelse(cad_data[,"CVA"] == "Y", 1, 0)
cad_data[,"Airway.disease"] = ifelse(cad_data[,"Airway.disease"] == "Y", 1, 0)
cad_data[,"Thyroid.Disease"] = ifelse(cad_data[,"Thyroid.Disease"] == "Y", 1, 0)
cad_data[,"CHF"] = ifelse(cad_data[,"CHF"] == "Y", 1, 0)
cad_data[,"DLP"] = ifelse(cad_data[,"DLP"] == "Y", 1, 0)
cad_data[,"Weak.Peripheral.Pulse"] = ifelse(cad_data[,"Weak.Peripheral.Pulse"] == "Y", 1, 0)
cad_data[,"Lung.rales"] = ifelse(cad_data[,"Lung.rales"] == "Y", 1, 0)
cad_data[,"Systolic.Murmur"] = ifelse(cad_data[,"Systolic.Murmur"] == "Y", 1, 0)
cad_data[,"Diastolic.Murmur"] = ifelse(cad_data[,"Diastolic.Murmur"] == "Y", 1, 0)
cad_data[,"Dyspnea"] = ifelse(cad_data[,"Dyspnea"] == "Y", 1, 0)
cad_data[,"Atypical"] = ifelse(cad_data[,"Atypical"] == "Y", 1, 0)
cad_data[,"Nonanginal"] = ifelse(cad_data[,"Nonanginal"] == "Y", 1, 0)
cad_data[,"Exertional.CP"] = ifelse(cad_data[,"Exertional.CP"] == "Y", 1, 0)
cad_data[,"LowTH.Ang"] = ifelse(cad_data[,"LowTH.Ang"] == "Y", 1, 0)
cad_data[,"LVH"] = ifelse(cad_data[,"LVH"] == "Y", 1, 0)
cad_data[,"Poor.R.Progression"] = ifelse(cad_data[,"Poor.R.Progression"] == "Y", 1, 0)

###################Change categorical variables to factors######################

#independent variables
#convert categorical to factors
cad_data[,"Sex"] = as.factor(cad_data[,"Sex"])
cad_data[,"Function.Class"] = as.factor(cad_data[,"Function.Class"])
cad_data[,"Region.RWMA"] = as.factor(cad_data[,"Region.RWMA"])
cad_data[,"VHD"] = as.factor(cad_data[,"VHD"])
cad_data[,"BBB"] = as.factor(cad_data[,"BBB"])


#convert binary variables to factors
cad_data[,"DM"] = as.factor(cad_data[,"DM"])
cad_data[,"HTN"] = as.factor(cad_data[,"HTN"])
cad_data[,"Current.Smoker"] = as.factor(cad_data[,"Current.Smoker"])
cad_data[,"EX-Smoker"] = as.factor(cad_data[,"EX-Smoker"])
cad_data[,"FH"] = as.factor(cad_data[,"FH"])
cad_data[,"Obesity"] = as.factor(cad_data[,"Obesity"])
cad_data[,"CRF"] = as.factor(cad_data[,"CRF"])
cad_data[,"CVA"] = as.factor(cad_data[,"CVA"])
cad_data[,"Airway.disease"] = as.factor(cad_data[,"Airway.disease"])
cad_data[,"Thyroid.Disease"] = as.factor(cad_data[,"Thyroid.Disease"])
cad_data[,"CHF"] = as.factor(cad_data[,"CHF"])
cad_data[,"DLP"] = as.factor(cad_data[,"DLP"])
cad_data[,"Edema"] = as.factor(cad_data[,"Edema"])
cad_data[,"Weak.Peripheral.Pulse"] = as.factor(cad_data[,"Weak.Peripheral.Pulse"])
cad_data[,"Lung.rales"] = as.factor(cad_data[,"Lung.rales"])
cad_data[,"Systolic.Murmur"] = as.factor(cad_data[,"Systolic.Murmur"])
cad_data[,"Diastolic.Murmur"] = as.factor(cad_data[,"Diastolic.Murmur"])
cad_data[,"Typical.Chest.Pain"] = as.factor(cad_data[,"Typical.Chest.Pain"])
cad_data[,"Dyspnea"] = as.factor(cad_data[,"Dyspnea"])
cad_data[,"Atypical"] = as.factor(cad_data[,"Atypical"])
cad_data[,"Nonanginal"] = as.factor(cad_data[,"Nonanginal"])
cad_data[,"Exertional.CP"] = as.factor(cad_data[,"Exertional.CP"])
cad_data[,"LowTH.Ang"] = as.factor(cad_data[,"LowTH.Ang"])
cad_data[,"Q.Wave"] = as.factor(cad_data[,"Q.Wave"])
cad_data[,"St.Elevation"] = as.factor(cad_data[,"St.Elevation"])
cad_data[,"St.Depression"] = as.factor(cad_data[,"St.Depression"])
cad_data[,"Tinversion"] = as.factor(cad_data[,"Tinversion"])
cad_data[,"LVH"] = as.factor(cad_data[,"LVH"])
cad_data[,"Poor.R.Progression"] = as.factor(cad_data[,"Poor.R.Progression"])

################################################################################

#convert dependent variable to factor
cad_data[,"Cath"] = as.factor(cad_data[,"Cath"])

################################################################################

#recheck data types
str(cad_data)

```

## 3.4. Summary statistics for numeric independent variables

```{r}
#################################summary data###################################

#get summary stats
summary_stats = t(data.frame(do.call(cbind, 
                                           lapply(cad_data, summary)
                                           )))
for (row in 1:nrow(summary_stats)) { #for variable in summary stats dataframe...
  if (is.numeric(cad_data[, row]) == TRUE) { #if variable is numeric ...
    summary_stats[row,] = round(as.numeric(summary_stats[row,]), 3) #round off
  }
  else { #else if variable is categorical jump to next variable
    next
  }
}

################################################################################
#report on summary stats of numeric variables

#empty vector
summary_stats_of_numeric = c()

#create data frame of summary stats of numeric variables
for (row in 1:nrow(summary_stats)) {  #for variable in summary stats data frame...
  if (is.numeric(cad_data[, row]) == TRUE) { #if variable is numeric ...
    
    #append column as row to dataframe
    summary_stats_of_numeric = rbind(summary_stats_of_numeric, summary_stats[row, ])
  }
}

#####################additional summary statistics##############################

#add sd and var columns to summary stats table
std = c()
variance = c()

##standard deviation
for (col in 1:ncol(cad_data)) {  #for each column in the cad data set...
  if (is.numeric(cad_data[, col]) == TRUE) { #if variable is numeric...
    std[col] = round(sd(cad_data[, col]), 3) #calc. the standard deviation...
  }
  else { #else if categorical...
    next #pass
    
  }
}

#remove NAs of categorical variables
std = na.omit(std)

################################################################################

##variance
for (col in 1:ncol(cad_data)) {   #for each column in the cad data set...
  if (is.numeric(cad_data[, col]) == TRUE) {   #if variable is numeric...
    variance[col] = round(var(cad_data[, col]), 3)  #calc. the variance
  }
  else { #else if categorical
    next #pass
    
  }
}


#remove NAs of categorical variables
variance = na.omit(variance)


##append standard deviation and variance to summary statistics data frame
summary_stats = cbind(summary_stats_of_numeric, variance, std)

################################################################################

#assign row names to numeric variables date frame
colnames_of_numeric_var = c()
for (col in 1:ncol(cad_data)) {   #for each column in the cad data set...
  if (is.numeric(cad_data[, col]) == TRUE) {   #if variable is numeric...
    
    #append variable name to list
   colnames_of_numeric_var = c(colnames_of_numeric_var, colnames(cad_data)[col])  
  }
  else { #else if categorical...
    next #pass
    
  }
}

#row names of dataframe
rownames(summary_stats) = colnames_of_numeric_var

#save table data
repmod::make_word_table(summary_stats, 
                        paste0(path_to_figures_tables, 
                               "\\summary_stats"), 
                        info = NULL, 
                        use.rownames = TRUE)

```


## 3.5. Box plots, bar plots and histograms showing distribution of observations for each variable

```{r}

num = 0 #counter

#Boxplot of numeric independent variables
par(mfrow=c(1, 3))
for (col in 1:ncol(cad_data)) {   #for each column in the cad data set...
  if (is.numeric(cad_data[, col]) == TRUE) {   #if variable is numeric...
    num = num + 1
    boxplot(cad_data[, col], main=paste0(num, ". ", colnames(cad_data)[col]),
            col="lightgreen") #plot boxplot
  }
  else { #else if categorical variable...
    next #pass
  }
}


################################################################################

#Histograms of numeric independent variables

#counter
numeric_counter = 0

#hisograms
par(mfrow=c(1, 3), pty="s")
for (col in 1:ncol(cad_data)) {   #for each column in the cad data set...
  if (is.numeric(cad_data[, col]) == TRUE) {   #if variable is numeric...
    numeric_counter = numeric_counter  + 1
    hist(cad_data[, col], main=paste0(numeric_counter, ". ", 
                                      colnames(cad_data)[col]),
         col="lightblue", xlab=colnames(cad_data)[col])  # plot histogram
  }
  else {  #else if categorical variable...
    next #pass
  }
}

#################################################################################

#Bar plots of categorical variables

#counter
categor_counter = 0

#bar plots
par(mfrow=c(1, 3), pty="s")
for (col in 1:ncol(cad_data)) {   #for each column in the cad data set...
  if (is.numeric(cad_data[, col]) == TRUE) {   #if variable is numeric...
    next #pass
  }
  else {  #else if categorical variable...
    
    #count of variable
    categor_counter = categor_counter + 1
    
    #table summary of counts for each category
    tab = table(cad_data[, col])
    
    #barplot
    bp = barplot(tab, 
            main=paste0(categor_counter, ". ", colnames(cad_data)[col]),
            col="pink", ylab="Frequency",
            ylim=c(0, max(tab)*1.3), xlab="class", las=2)
    text(bp, tab, labels=tab, pos=3)

  }
}


```




## 3.6. Correlation

```{r}

#generate correlation matrix

##empty vector
numeric_variables = c()  

for (col in 1:ncol(cad_data)) {   #for each column in the cad data set...
  if (is.numeric(cad_data[, col]) == TRUE) {   #if variable is numeric...
    
    #append numeric variable to data frame
    numeric_variables = cbind(numeric_variables, cad_data[,col])
  }
  else { #else if categorical variable...
    next #pass
  }
}

#add column names to data frame
colnames(numeric_variables) = c("Age", "Weight", "Length", "BMI",
                               "BP", "PR", "FBS", "CR", "TG", "LDL", "HDL", "BUN", 
                               "ESR", "HB", "K", "Na", "WBC", "Lymph",  "Neut", 
                               "PLT", "EF-TTE")

#save image
png(paste0(path_to_figures_tables, "\\cor_plot.png"))

#correlation matrix
cor_matrix_of_numeric_var = cor(numeric_variables,  method="pearson") 

#Correlation plot
(corrplot::corrplot(cor_matrix_of_numeric_var, tl.cex=0.9, type="upper")) 

dev.off()

################################################################################

#Which redundant variables should be removed?
caret::findCorrelation(cor(numeric_variables), cutoff=0.7) 

#remove "Weight" and "Lymph"
cad_data = cad_data %>% subset(select=-c(Weight, Lymph))

#number of indep variables left
ncol(cad_data)-1

```

## 3.7 Removing constant categorical/binary independent variables

```{r}
#also remove Exertional.CP because all records in 1 class
cad_data = cad_data %>% subset(select=-c(Exertional.CP)) 

#number of indep variables left
ncol(cad_data)-1

```

\newpage

# 4. Discretization

```{r}

#CR
cad_data$CR = as.factor(ifelse(cad_data$CR < 0.7, "Low", ifelse(cad_data$CR >=0.7 & cad_data$CR<=1.5, "Normal", "High")))

#FBS
cad_data$FBS = as.factor(ifelse(cad_data$FBS < 70, "Low", ifelse(cad_data$FBS >=70 & cad_data$FBS<=105, "Normal", "High")))

#LDL
cad_data$LDL = as.factor(ifelse(cad_data$LDL <= 130, "Normal", "High"))

#HDL
cad_data$HDL = as.factor(ifelse(cad_data$HDL < 35, "Low", "Normal"))

#BUN
cad_data$BUN = as.factor(ifelse(cad_data$BUN < 7, "Low", ifelse(cad_data$BUN >=7 & cad_data$BUN<=20, "Normal", "High")))

#ESR
cad_data$ESR = as.factor(ifelse(cad_data$Sex == "Male" & cad_data$ESR <= cad_data$Age/2, "Normal", 
                  ifelse(cad_data$Sex == "Fmale" & cad_data$ESR <= cad_data$Age/2 + 5, "Normal", 
                         ifelse(cad_data$Sex == "Male" & cad_data$ESR > cad_data$Age/2, "High", 
                                ifelse(cad_data$Sex == "Fmale" & cad_data$ESR > cad_data$Age/2 + 5, "High", "")))))


#HB
cad_data$HB = as.factor(ifelse(cad_data$Sex == "Male" & cad_data$HB < 14, "Low", 
                  ifelse(cad_data$Sex == "Fmale" & cad_data$HB < 12.5, "Low", 
                         ifelse(cad_data$Sex == "Male" & cad_data$HB >= 14 & cad_data$HB <= 17 , "Normal", 
                                ifelse(cad_data$Sex == "Fmale" & cad_data$HB >= 12.5 & cad_data$HB <= 15 , "Normal",
                                
                                ifelse(cad_data$Sex == "Male" & cad_data$HB > 17 , "High", 
                                ifelse(cad_data$Sex == "Fmale" & cad_data$HB > 15  , "High", "")))))))


#K
cad_data$K = as.factor(ifelse(cad_data$K < 3.8, "Low", ifelse(cad_data$K >=3.8 & cad_data$K<=5.6, "Normal", "High")))

#Na
cad_data$Na = as.factor(ifelse(cad_data$Na < 136, "Low", ifelse(cad_data$Na >=136 & cad_data$Na<=146, "Normal", "High")))

#WBC
cad_data$WBC = as.factor(ifelse(cad_data$WBC < 4000, "Low", ifelse(cad_data$WBC >=4000 & cad_data$WBC<=11000, "Normal", "High")))

#PLT
cad_data$PLT = as.factor(ifelse(cad_data$PLT < 150, "Low", ifelse(cad_data$PLT >=150 & cad_data$PLT <=450, "Normal", "High")))

#EF
cad_data$`EF-TTE` = as.factor(ifelse(cad_data$`EF-TTE` <= 50, "Low","Normal"))

#Region with RWMA
cad_data$Region.RWMA = as.factor(ifelse(cad_data$Region.RWMA == 0, "Normal","High"))

#Age
cad_data$Age = as.factor(ifelse(cad_data$Sex == "Male" & cad_data$Age <= 45, "Normal", 
                  ifelse(cad_data$Sex == "Fmale" & cad_data$Age <= 55, "Normal", 
                         ifelse(cad_data$Sex == "Male" & cad_data$Age > 45, "High", 
                                ifelse(cad_data$Sex == "Fmale" & cad_data$Age > 55, "High",
                                       "")))))


#BP
cad_data$BP = as.factor(ifelse(cad_data$BP < 90, "Low", ifelse(cad_data$BP >=90 & cad_data$BP <=140, "Normal", "High")))

#Pulse rate
cad_data$PR = as.factor(ifelse(cad_data$PR < 60, "Low", ifelse(cad_data$PR >=60 & cad_data$PR <=100, "Normal", "High")))

#TG
cad_data$TG = as.factor(ifelse(cad_data$TG <= 200, "Normal", "High"))

#Function class
cad_data$Function.Class =  as.factor(ifelse(cad_data$Function.Class == 1, "Normal", "High"))

################################################################################

#Neut

cad_data$Neut =  as.factor(ifelse(cad_data$Neut < 40, "Low",
                                            ifelse(cad_data$Neut >= 40 & cad_data$Neut <= 60, "Normal", "High")))
                                                   

################################################################################

#BMI

cad_data$BMI = as.factor(ifelse(cad_data$BMI < 18.5, "Underweight", 
                                ifelse(cad_data$BMI >=18.5 & cad_data$BMI<25, "Healthy", 
                                       ifelse(cad_data$BMI >= 25 & cad_data$BMI < 30, "Overweight",
                                              "Obesity"))))

# If your BMI is less than 18.5, it falls within the underweight range.
# If your BMI is 18.5 to <25, it falls within the healthy weight range.
# If your BMI is 25.0 to <30, it falls within the overweight range.
# If your BMI is 30.0 or higher, it falls within the obesity range.

################################################################################

#Length

cad_data$Length = as.factor(ifelse(cad_data$Length  < 160, "Length < 160 ", 
                                ifelse(cad_data$Length  >=160 & cad_data$Length <180, "160 ≤ Length < 180",
                                       "Length ≥ 180")))

```


\newpage

# 5. Stratified sampling

```{r}


## Change column names to be more readable
colnames(cad_data)[8] = "Ex.Smoker"
colnames(cad_data)[50] = "EF"


##class imbalance observed
table(cad_data$Cath)

##ensure balanced data set by downsampling majority class of dependent variable
set.seed(123)
cad_data <- data.frame(downSample(x = cad_data[, -which(names(cad_data) == "Cath")], 
                                  y = cad_data$Cath))
colnames(cad_data)[53] = "Cath"

##check that classes are balanced
cad_data$Cath  %>% summary()

```


\newpage

# 6. Feature selection

```{r}

#variable selection using svm

##empty vector
df =c()

##convert x-data to numeric
for (col in 1:ncol(cad_data)) {
  df = cbind(df, as.numeric(cad_data[,col]))
}

##convert data frame to data frame object
df = data.frame(df)

##remove numeric y-data
df = df %>% subset(select=-c(X53))

##append factor y-data
df = cbind(df, cad_data[, "Cath"])

##set column names to df
colnames(df) = colnames(cad_data)

##fit svm model
set.seed(123) #set seed to get the same results
fit = svm(Cath ~ ., df, scale=F,type = 'C-classification',
                 kernel = 'linear')

## get model coefficients
coefs = coef(fit)

##get variables  where coefficients < 0.6
svm.rm = c()
svm.rm =  c(svm.rm, coefs[abs(coefs) < 0.6])
svm.rm = names(svm.rm)

################################################################################

#variable selection using random Forest

##fit random forest model
set.seed(123) #set seed to get the same results
fit = randomForest(Cath ~ ., cad_data, scale=F)

##get variable importance
varimps = varImp(fit, scale=F)

##convert varimps to dataframe
imp <- data.frame(varimps)
imp <- data.frame(overall = imp$Overall, names = rownames(imp))

##get variables where variable importance < 0.6
rf.rm = imp$names[imp$overall < 0.6]

#################################################################################

#variable selection using ROC

##get variable importance
set.seed(123) #set seed to get the same results
roc_imp = filterVarImp(x = cad_data[,1:52], y=cad_data$Cath)

##get variables to be removed
roc.rm = rownames(roc_imp)[roc_imp$Cad < 0.6]


################################################################################

#variable selection using tree

##fit tree model
set.seed(123) #set seed to get the same results
fit = rpart(Cath ~ ., cad_data)

## get variable importance
varimps = data.frame(varImp(fit, scale=F))

##get variables to be removed
tree.rm = rownames(varimps)[varimps$Overall < 0.6]

################################################################################

#Find variables to discard

intersect1 = intersect(roc.rm, rf.rm)
intersect2 = intersect(intersect1, svm.rm)
intersect3 = intersect(intersect2, tree.rm)

################################################################################
# •	"Ex.Smoker" (removed from dataset) *
# •	"CRF" (removed from dataset)        *   
# •	 "CVA" (removed from dataset)        *      
# •	"Airway.disease" (removed from dataset) *
# •	“Thyroid.Disease” (removed from dataset)   *
# •	"CHF" (Incorrectly identified as unimportant due to class imbalance)  *                 
# •	"PR" (Incorrectly identified as unimportant due to data sparsity)      *                            
# •	"Weak.Peripheral.Pulse" (Incorrectly identified as unimportant due to class imbalance) *                   
# •	 "Lung.rales"    (Incorrectly identified as unimportant due to class imbalance) *
# •	"LVH" (removed from dataset)   *
# •	"Poor.R.Progression" (removed from dataset)   *
# •	"BBB" (removed from dataset)   *
# •	"Na" (removed from dataset)   *
# •	"PLT" (removed from dataset)   *

##################################################################################

## discard variables that were found to be unimportant, only if they do no 
## occur as a variable under the "symptom and examination" group of variables
cad_data = subset(cad_data, select=-c(Ex.Smoker, CRF, CVA, Airway.disease, 
                                      Thyroid.Disease, LVH, Poor.R.Progression, 
                                      BBB, Na, PLT))

## also discard variables that were discarded by:
## https://www.sciencedirect.com/science/article/pii/S0169260713000801?via%3Dihub

cad_data = cad_data %>% subset(select=-c(Atypical, Nonanginal, FBS, 
                                         Diastolic.Murmur, Current.Smoker)) *

##remaining independent variables
ncol(cad_data)-1
colnames(cad_data)

```

\newpage

# 7. Convert multinomial features to binomial features

Rule:
If feature f is polynomial, change it to binomial by mapping the values having direct relationship to CAD, to 1 and others to 0. For example HB has three values: Low, Normal, and High. “Low” and “High” are mapped to the "1" group, and “Normal” is mapped to the "0" group. 

```{r}

#check data types
str(cad_data)

#################################################################################

#Convert multinomial features to binomial features:
        ##Rule: Normal, average and median classes are coded as 0. 
        ##Extremes such as low and high classes are coded as 1 since
        ##it has an effect on CAD

##################################################################################

#Length with 3 levels
cad_data$Length = as.factor(ifelse(cad_data$Length  == "160 ≤ Length < 180", 0, 1))

####################################################################################

#BMI with 4 levels
cad_data$BMI = as.factor(ifelse(cad_data$BMI == "Healthy", 0, 1))

####################################################################################

#PR with 3 levels
cad_data$PR = as.factor( ifelse(cad_data$PR == "Normal", 0, 1))

####################################################################################

#CR with 3 levels
cad_data$CR = as.factor(ifelse(cad_data$CR == "Normal", 0, 1))

####################################################################################

#BUN with 3 levels
cad_data$BUN = as.factor(ifelse(cad_data$BUN == "Normal", 0, 1))

####################################################################################

#HB with 3 levels
cad_data$HB = as.factor(ifelse(cad_data$HB == "Normal", 0, 1))

####################################################################################

#K with 3 levels
cad_data$K = as.factor(ifelse(cad_data$K == "Normal", 0, 1))

####################################################################################

#WBC with 3 levels
cad_data$WBC = as.factor(ifelse(cad_data$WBC == "Normal", 0, 1))

####################################################################################

#Neut with 3 levels
cad_data$Neut = as.factor(ifelse(cad_data$Neut == "Normal", 0, 1))

####################################################################################

#VHD with 4 levels
cad_data$VHD = as.factor(ifelse(cad_data$VHD == "N", 0, 1))

#####################################################################################

#BP with levels
cad_data$BP = as.factor(ifelse(cad_data$BP == "Normal", 0, 1))

#####################################################################################


#BP with levels
cad_data$Function.Class = as.factor(ifelse(cad_data$Function.Class == "Normal", 
                                           0, 1))

#####################################################################################

#TG with 2 levels
cad_data$TG = as.factor(ifelse(cad_data$TG == "Normal", 0, 1))

#####################################################################################

#LDL with 2 levels
cad_data$LDL = as.factor(ifelse(cad_data$LDL == "Normal", 0, 1))

#####################################################################################

#HDL with 2 levels
cad_data$HDL = as.factor(ifelse(cad_data$HDL == "Normal", 0, 1))

#####################################################################################

#ESR with 2 levels
cad_data$ESR = as.factor(ifelse(cad_data$ESR == "Normal", 0, 1))

#####################################################################################

#EF with 2 levels
cad_data$EF = as.factor(ifelse(cad_data$EF == "Normal", 0, 1))

#####################################################################################

#EF with 2 levels
cad_data$Region.RWMA = as.factor(ifelse(cad_data$Region.RWMA == "Normal", 0, 1))

#####################################################################################

#Age with 2 levels
cad_data$Age = as.factor(ifelse(cad_data$Age == "Normal", 0, 1))

#####################################################################################

#Sex with 2 levels. Female = 1 and male = 0
cad_data$Sex = as.factor(ifelse(cad_data$Sex == "Fmale", 1, 0))

#####################################################################################

#recheck data types to see that they are binomial
str(cad_data)

```


\newpage

# 8. Association rule mining

## i. Conversion of data to transaction data

```{r}
#recheck dim
dim(cad_data)

#convert data frame to transaction data frame
trans <- as(cad_data, "transactions")

```


## ii. Exploratory data analysis on transaction data

```{r}

#inspect first 5 items
set.seed = 123
inspect(trans[1:5])

#Checking the frequency of the first 5 items in transaction database
set.seed = 123
itemFrequency(trans[,1:5])

#Frequency plot with items having a frequency of at  least 70%
set.seed = 123
itemFrequencyPlot(trans, support = 0.7, col=rainbow(length(trans)))


#maximum frequency for the top 25 records
set.seed = 123
itemFrequencyPlot(trans, topN=25, type = "relative", col=rainbow(length(trans)))

#Below is the visualization of tthe Sparse Matrix for first 25 transactions.
set.seed = 123
image(trans[1:25])

```


## 8.1 Objective 1: Use association rule mining methods to determine the features that are mostly associated with CAD.

```{r}

## Set RHS Cath = Cad 

### ensure a maximum rule length of 4, so that MAX length of LHS = 3 and length of RHS = 1

set.seed = 123
rules<-apriori(data=trans, maxlen=4, parameter=list(supp=0.01,conf = 0.8), appearance =list(default="lhs",rhs="Cath=Cad"), control =list(verbose=F)) 

#check number of rules
rules

# get redundant rules into a vector
set.seed = 123
redundant <- which (colSums (is.subset (rules, rules)) > 1)

# remove redundant rules
rules <- rules[-redundant]
inspect(rules[1:5])

#number of rules
rules

#summary of rules
summary(rules)

# sort rules by support, confidence, and lift
rules<-sort(rules,decreasing=TRUE,by="support")
inspect(rules[1:5])

#write rules to file
write.csv(as(rules, "data.frame"), paste0(path_to_figures_tables, 
                                                            "\\objective1_rules_sorted_by_support.csv"))


#plot network
plot(head(rules, 5), method="graph",control=list(type="items",main=""))


```

### 8.1.1 Scatter plot of interesting rules

```{r, echo=FALSE, warnings=FALSE,fig.width=8, fig.height=6}

plot(rules,main="")  

#Rules with high lift typically have low support. 
#The most interesting rules reside where the confidence is high and support 
#is low i.e., having large lift values

```

### 8.1.2 Plot graph-based visualisation

```{r}

#sort by confidence
subrules <- head(sort(rules, by="confidence"), 5) #change head value
inspect( subrules[1:5])


#plot network
plot(subrules, method="graph",control=list(type="items",main=""))



```


### 8.1.3 Grouped plot for all rules

```{r}

plot(rules, method="grouped") 

```

### 8.1.4 Interactive plot for all rules

```{r}
plot(rules, method = "graph", measure = "lift", shading = "confidence", engine = "htmlwidget")
```

### 8.1.5 Final most interesting rules

```{r}

# sort rules by lift
rules_ordered_by_lift<-sort(rules,decreasing=TRUE,by="lift")  
inspect(rules_ordered_by_lift[1:5])

#write rules to file
write.csv(as(rules_ordered_by_lift, "data.frame"), paste0(path_to_figures_tables, 
                                                            "\\objective1_final_rules.csv"))

# sort rules by confidence
rules_ordered_by_confidence<-sort(rules,decreasing=TRUE,by="confidence")

#assess by confidence
inspect(rules_ordered_by_confidence[1:5])

```

\newpage

## 8.2 Objective 2: Find any other interesting association rules.
* Rules with high lift typically have low support.
* The most interesting rules reside on the support/confidence border which can be clearly seen by a scatter plot.


* Based on output from objective 1 above, DM is a highly associated with CAD. 
* From theory, we know that diabetic individuals are at high risk of CAD. 
* It is worth investigating the factors that are associated with DM.

```{r apriori, fig.show= "hide", results = "hide"}

##decrease minsup and increase confidence
## keep increasing minlen until interesting associations cease to exist
## choose minlen = 6 because minlen = 7 generates 0 rules

set.seed = 123
rules<-apriori(data=trans,parameter=list(supp=0.001,conf =0.9), 
               control =list(verbose=F), target = "rules", 
                appearance = list(default="lhs", rhs="DM=1"), minlen=6)

####################################################################################

#number of rules
rules #554258 rules

#view rules
inspect(rules[1:5])

#summary of rules
summary(rules)

####################################################################################

# sort rules by lift
rules_ordered_by_lift<-sort(rules,decreasing=TRUE,by="lift") 

#show first few rules
inspect(rules_ordered_by_lift[1:5])

#save rules to csv
##uncomment if required
#write.csv(as(inspect(rules_ordered_by_lift[1:25]), "data.frame"), paste0(path_to_figures_tables,"\\objective2_final_rules.csv"))


####################################################################################

```


### 8.2.1 Scatter plot of interesting rules

```{r, echo=FALSE, warnings=FALSE,fig.width=8, fig.height=6}

##Rules with high lift typically have low support. 
##The most interesting rules reside where the confidence is high and support is low i.e., having large lift values


##plot rules
plot(rules, measure=c("support", "confidence"), shading="lift",control = list(col=rainbow(7))) 

```


### 8.2.2 Plot graph-based visualisation

```{r}

#sort by lift
subrules2 <- head(sort(rules, by="confidence"), 5)
inspect( subrules2[1:5])

#plot network
plot(subrules2, method="graph",control=list(type="items",main=""), lwd=4)


```

